Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models

### 2025-12-19 Ecov_study/mortality/manuscript/manuscript_R1.Rmd
revision for project 0 complete, save work for M-Ecov revision

Our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20; @stockmiller21; @milleretal_inreview]. The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies [@stocketal21;@legaultetal23; @lietal24; @brittenetal_inreview; @lietal_inreview_a] and is used to assess many commercially important stocks in the Northeast U.S. [e.g., @nefsc22;@nefsc22a;@nefsc24]. We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) to generate all results.

@@

For each simulated data set we fit a set of 12 estimating models (EMs). The factors that distinguish the estimating models, also described in detail below, include source of population process error type (3 levels) whether (median) $M$ was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on $M$ was estimated or not (2 levels).

@@

>>>>

Our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and estimation models (EMs) [@millerstock20; @stockmiller21; @milleretal_inreview]. The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies [@stocketal21;@legaultetal23; @lietal24; @brittenetal_inreview; @lietal_inreview_a] and is used to assess many commercially important stocks in the Northeast U.S. [e.g., @nefsc22;@nefsc22a;@nefsc24]. We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) to generate all results.

@@

For each simulated data set we fit a set of 12 EMs. The factors that distinguish the estimating models, also described in detail below, include source of population process error type (3 levels) whether (median) $M$ was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on $M$ was estimated or not (2 levels).

@@

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics[width = 1.4\textwidth]{convergence_classification_plots}
\end{center}
\caption{Classification trees indicating primary factors determining convergence as defined by providing Hessian-based standard errors for R, R+S, and R+M OMs. Nodes denote percent convergence (top) and number of fits (bottom)  for the corresponding subset. Lower or higher convergence rates are indicated by more red or green polygons, respectively}\label{conv_class}
\end{figure}
\end{landscape}

\begin{table}
\caption{For each OM process error source (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence (providing Hessian-based standard errors) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
{\input{convergence_PRD_table}}
\end{table}

\begin{table}
\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq. \ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_converged_PRD_table}
{\input{bias_mean_M_converged_PRD_table}}
\end{table}

\begin{table}
\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq. \ref{bias_regression_response} for the the standard error of the MLE for median natural mortality rate ($\widehat {SE}\left(\widehat \beta_M\right)$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_se_PRD_table}
{\input{bias_mean_M_se_PRD_table}}
\end{table}

\begin{table}
\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq. \ref{bias_regression_response} for the covariate effect on natural mortality ($\beta_E$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_ecov_beta_converged_PRD_table}
{\input{bias_ecov_beta_converged_PRD_table}}
\end{table}

\begin{table}
\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq. \ref{bias_regression_response} for the the standard error of the MLE for covariate effect on natural mortality  ($\widehat {SE}\left(\widehat \beta_E\right)$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_se_PRD_table}
{\input{bias_ecov_beta_se_PRD_table}}
\end{table}

\begin{table}
\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq. \ref{bias_regression_response} for the terminal year SSB with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
{\input{bias_SSB_PRD_table}}
\end{table}

\begin{table}
\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq. \ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_complete_PRD_table}
{\input{bias_mean_M_complete_PRD_table}}
\end{table}

\begin{table}
\caption{For each OM process error source (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq. \ref{bias_regression_response} for the covariate effect on natural mortality ($\beta_E$) with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_ecov_beta_complete_PRD_table}
{\input{bias_ecov_beta_complete_PRD_table}}
\end{table}
